\documentclass{acm_proc_article-sp}
\begin{document}

\title{Enhancing Undergraduate Operating Systems Course Delivery Through 
A Real Operating System}

\numberofauthors{2}
\author{
\alignauthor
Joseph Anthony C. Hermocilla\\
       \affaddr{Institute of Computer Science}\\
       \affaddr{University of the Philippines Los Ba\~nos}\\
       \affaddr{College 4031, Laguna, Philippines}\\
       \email{jachermocilla@uplb.edu.ph}
\alignauthor
Joseph Emmanuel DL. Dayo\\
       \affaddr{Friendster, Incorporated}\\      
       \email{joseph.dayo@gmail.com}
}
\date{02 September 2009}

\maketitle
\begin{abstract}
Operating systems is a fundamental knowledge area in computer science education
emphasized by the ACM and IEEE review task force for the computer science 
curriculum. The many concepts, services, and functions supported by an operating
system makes it a complex piece of software composed of hundreds to thousands
lines of source code. This paper presents the experiences and results of using 
ICS-OS to teach an undergraduate operating systems course. ICS-OS a fork of
DEX-OS, is a fully functional operating system with a small source 
code base which makes it ideal for teaching. Traditional teaching approaches
has relied on simulations and user space application development. However, 
these are not enough since students do not see their programs working at the
kernel level. At the end of the semester the course was offered, the students 
were able to create interesting projects that demonstrate their understanding
and appreciation.
\end{abstract}

\keywords{operating systems, computer science education, information technology
 education} 

\section{Introduction}
An operating system is a piece of software that provides a layer 
of abstraction to make it easy and convenient for users to control the hardware.

\section{Significance}
Current approaches to teaching operating systems to undergraduates does not
involve programming the components of an actual operating system that can run
on real hardware. Simulations are used or user level application development.
Students, however, are more interested in writing code that can run at the
kernel level instead of just user level. 

\section{Related Work}

\subsection{Minix}
Minix has been around for several years already and has been the running 
example in the textbook, Operating Systems: Design and Implementation by 
Andrew Tanenbaum. Since its initial release, Minix has grown in size in terms
of source code as well as in complexity. 

\subsection{Nachos}
Nachos is instructional software for teaching undergraduate, and potentially
graduate, level operating systems courses. The Nachos distribution
comes with: 

   an overview paper
   simple baseline code for a working operating system
   a simulator for a generic personal computer/workstation
   sample assignments
   a C++ primer (Nachos is written in an easy-to-learn subset of C++, 
     and the primer helps teach C programmers our subset)

The assignments illustrate and explore all areas of modern operating
systems, including threads and concurrency, multiprogramming, 
system calls, virtual memory, software-loaded TLB's, file systems, 
network protocols, remote procedure call, and distributed systems.


\subsection{GeekOS}

GeekOS is another instructional operating system that runs on real hardware. 
It was developed by David Hovemeyer and used for instruction at the Department 
of Computer Science, University of Maryland. The design goals for GeekOS are
simplicity, realism, and understandability. The main features of GeekOS include interrupt handling, heap memory allocator, time-sliced kernel threads with 
static priority scheduling, mutexes and condition variables, user mode 
with segmentation-based memory protection, and device drivers for keyboard 
and VGA.


\subsection{Dex-OS}
Dex-OS was developed by Joseph Emmanuel Dayo. Dex-OS was opensourced under 
the GPL with the original project hosted at SourceForge. Its design is based on
an Aspect Oriented Approach to address cross-cutting concerns in operating 
systems design and implementation. A layered approach was taken based on 
modern operating systems like linux and windows. The following are the main
components of DEX-OS.

\textit{Memory Manager}

The memory manager is responsible for allocating memory to various components
of the operating system, including processes. It is further divided into 
low-level memory manager for setting up the GDT and high-level memory manager
for high-level functions such as malloc() and free().

\textit{Device Manager}

The device manager handles operations related to common devices and peripherals.
Device drivers for keyboard, display, and disk drives are managed by this 
module. Each device being managed is identified by a device id. 

\textit{Process Manager}

The process manager is responsible for handling process management functions
including process creation, process dispatching, process scheduling, process
synchronization, and process termination. 

\textit{Virtual Filesystem}

The Virtual File System (VFS) provides an device-independent abstraction for
various directory and file operations. Functions that are part of the VFS are
implemented by the low-level file system drivers. In particular, DEX-OS 
supports the Microsoft FAT12, FAT16, and FAT32 file systems.

\section{Methodology}

\subsection{Setup a linux development environment}
The original development environment for DEX-OS was in Windows and uses the 
Dev-C++ IDE. Since the laboratory rooms of the students were classes are held
use linux, the first step was to migrate the development environment to 
linux. The following steps were taken.

The Makefile is the primary mechanism to build the ICS-OS kernel and 
generate a distribution floppy. The original Makefile and Linker Script
was converted to generate an ELF kernel image (linux) instead of COFF.


\section{Results and Discussion}

\section{Conclusion}

\bibliographystyle{abbrv}
\bibliography{jach-ncite09.bbl}
\balancecolumns


\end{document}
