%
%
\documentclass{article}
\title{Paper Review\\
	Multiagent Systems: A Survey from a Machine Learning Perspective\cite{pstone00}}
\author{Joseph Anthony C. Hermocilla}

\begin{document}
\maketitle
\section{Summary}
Multiagent systems (MAS) is a relatively new area of research under Distributed 
Artificial Intelligence (DAI). The survey paper gave an introduction to MAS
research by providing a taxonomy of MAS and discussing the issues and techniques
associated with each classification from a machine learning perspective. 
The taxonomy presented is based on the 
degree of heterogeneity and degree of communication, rather than on the domain. 
The paper also described an example domain, Predator and Prey, which was used in the  
discussion of issues and techniques for each classification. 


\section{Evaluation}
The survey paper is a good introduction to the field of Multiagent Systems research.
An individual new in the field can easily follow the flow of the discussion. The 
presentation of issues and techniques encourages the reader to think of other ideas
to improve existing approaches. Reference to existing and related works have been
provided for futher exploration of the field.

\section{Syntesis}
Multiagent systems research seems to be a promising area of research. Its distributed
nature allows computations to be performed in parallel at the individual agent level.  
Each agent in a MAS can pursue its own goals but eventually the achievement of the
individual(local) goal will contribute to the global goal.

\bibliographystyle{plain}
\bibliography{references}
\nocite{pfong04}

\end{document}

